---
title: "STA510 - Mandatory Assignment 1"
author: "Tommaso Facchin"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
geometry: margin=2.5cm
fontsize: 11pt
lang: en
---

```{r setup, include=FALSE}
rm(list = ls())

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```


# Problem 1

## a)


\[
\int_a^{a+20} c(x-a)(a+20-x)\,dx = 1.
\]

Using the substitution

\[
u = x-a,
\qquad
dx = du,
\]

we obtain

\[
c\int_0^{20}u(20-u)\,du = 1.
\]

\[
c\left[10u^2-\frac{u^3}{3}\right]_0^{20} = 1.
\]

\[
c\frac{4000}{3} = 1.
\]

\[
\boxed{c=\frac{3}{4000}}.
\]

The valid pdf is

\[
f(x;a)=
\begin{cases}
\displaystyle \frac{3}{4000}(x-a)(a+20-x),
& a \leq x \leq a+20,\\[6pt]
0, & \text{otherwise}.
\end{cases}
\]


## b) 

\[
E[X]
=
\int_a^{a+20}
x \cdot c(x-a)(a+20-x)\,dx.
\]

I used the substitution like before and got:

\[
E[X]
=
c\int_0^{20}
\left(
20u^2-u^3+20au-au^2
\right)\,du.
\]

\[
E[X]
=
c
\left[
\frac{20u^3}{3}
-\frac{u^4}{4}
+10au^2
-\frac{au^3}{3}
\right]_0^{20}.
\]

\[
\boxed{E[X]=a+10}.
\]


---

\[
\operatorname{Var}(X)=E[X^2]-(E[X])^2.
\]

\[
E[X^2]
=
\int_a^{a+20}x^2f(x;a)\,dx.
\]

\[
E[X^2]
=
a^2+40a+400.
\]


\[
\operatorname{Var}(X)
=
(a^2+40a+400)-(a+10)^2.
\]

\[
\boxed{\operatorname{Var}(X)=100}.
\]

---

E[X] = a+10 means that a + 10m/s is the expected daily maximum wind speed during autumn. right in the middle of the minimum and maximum threshold.+

Var[X] = 100 means that the daily maximum wind speeds have a spread around their expected value with standard deviation 10m/s.




## c)

The cumulative distribution function is 

\[
F(x;a)
=
\int_a^x \frac{3}{4000}(t-a)(a+20-t)\,dt.
\]

using substitution: 

\[
F(x;a)
=
\frac{3}{4000}
\int_0^{x-a}u(20-u)\,du.
\]


\[
F(x;a)
=
\frac{3}{4000}
\left[
10u^2-\frac{u^3}{3}
\right]_0^{x-a}.
\]

\[
F(x;a)
=
\frac{3}{4000}
\left(
10(x-a)^2-\frac{(x-a)^3}{3}
\right).
\]


\[
\boxed{
F(x;a)
=
\frac{3}{4000}
\left(
10(x-a)^2-\frac{(x-a)^3}{3}
\right),
\qquad a\leq x\leq a+20.
}
\]


## d)



```{r}
a <- 5
n <- 5000

F <- function(x, a) {(3 / 4000) * (10 * (x - a)^2 - (x - a)^3 / 3)}

F_inv <- function(p, a) {
  uniroot(
    function(x) F(x, a) - p,
    lower = a,
    upper = a + 20
  )$root
}

set.seed(42)

u <- runif(n)

wind_speed <- sapply(u, F_inv, a = a)

```


```{r}
hist(
  wind_speed,
  breaks = "Scott",
  probability = TRUE,
  col = "lightblue",
  main = "Histogram of simulated wind speeds",
  xlab = "Wind speed (m/s)"
)

curve(
  (3 / 4000) * (x - 5) * (25 - x),
  from = 5,
  to = 25,
  add = TRUE,
  col = "red",
  lwd = 2
)
```

The simulated values follow the expected parabolic shape. 
The density is close to zero near the limits (\(5\) m/s and \(25\) m/s), and it reaches its maximum around \(15\) m/s. 


```{r ecdf}
plot(
  ecdf(wind_speed),
  main = "Empirical and true CDF",
  xlab = "Wind speed (m/s)",
  ylab = "Cumulative probability",
  col = "lightblue",
  lwd = 2
)

curve(
  (3 / 4000) * (10 * (x - 5)^2 - (x - 5)^3 / 3),
  from = 5,
  to = 25,
  add = TRUE,
  col = "red",
  lwd = 2
)
```

The empirical CDF is very similar to the true CDF. The empirical CDF is
a step function based on the simulated sample, whereas the true CDF is smooth.
The two curves are almost overlapping.
