---
title: "STA510 - Mandatory Assignment 1"
author: "Tommaso Facchin"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
geometry: margin=2.5cm
fontsize: 11pt
lang: en
---

```{r setup, include=FALSE}
rm(list = ls())

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```


# Problem 1

## d)

```{r}
a <- 5
n <- 5000

F <- function(x, a) {(3 / 4000) * (10 * (x - a)^2 - (x - a)^3 / 3)}

F_inv <- function(p, a) {
  uniroot(
    function(x) F(x, a) - p,
    lower = a,
    upper = a + 20
  )$root
}

set.seed(42)

u <- runif(n)

wind_speed <- sapply(u, F_inv, a = a)

```


```{r}
hist(
  wind_speed,
  breaks = "Scott",
  probability = TRUE,
  col = "lightblue",
  main = "Histogram of simulated wind speeds",
  xlab = "Wind speed (m/s)"
)

curve(
  (3 / 4000) * (x - 5) * (25 - x),
  from = 5,
  to = 25,
  add = TRUE,
  col = "red",
  lwd = 2
)
```

The simulated values follow the expected parabolic shape. 
The density is close to zero near the limits (\(5\) m/s and \(25\) m/s), and it reaches its maximum around \(15\) m/s. 


```{r ecdf}
plot(
  ecdf(wind_speed),
  main = "Empirical and true CDF",
  xlab = "Wind speed (m/s)",
  ylab = "Cumulative probability",
  col = "lightblue",
  lwd = 2
)

curve(
  (3 / 4000) * (10 * (x - 5)^2 - (x - 5)^3 / 3),
  from = 5,
  to = 25,
  add = TRUE,
  col = "red",
  lwd = 2
)
```

The empirical CDF is very similar to the true CDF. The empirical CDF is
a step function based on the simulated sample, whereas the true CDF is smooth.
The two curves are almost overlapping.


## e)

### iii) 

The theoretical acceptance rate is


\[
\frac{1}{M}
=
\frac{1}{3/2}
=
\boxed{\frac{2}{3}}
\approx0.6667.
\]

Approximately 66.67\% of the proposed values are expected to be
accepted.

To generate 5000 accepted samples, the expected number of proposals is

\[
\frac{5000}{2/3}
=
\boxed{7500}.
\]


## f)

```{r}
a <- 5
n <- 5000
M <- 3 / 2

wind_ar <- numeric(n)
proposed <- 0
i <- 1

while (i <= n) {
  
  x <- runif(1, min = a, max = a + 20)
  
  u <- runif(1)
  
  acceptance_probability <- (
    (3 / 4000) * (x - a) * (a + 20 - x)
  ) / (M * (1 / 20))
  
  proposed <- proposed + 1
  
  if (u <= acceptance_probability) {
    wind_ar[i] <- x
    i <- i + 1
  }
}

empirical_rate <- n / proposed
empirical_rate
```


### i)

```{r}
hist(
  wind_ar,
  breaks = "Scott",
  probability = TRUE,
  col = "lightblue",
  main = "AR simulated wind speeds",
  xlab = "Wind speed (m/s)",
  ylab = "Density"
)

curve(
  (3 / 4000) * (x - 5) * (25 - x),
  from = 5,
  to = 25,
  add = TRUE,
  col = "red",
  lwd = 2
)
```


The histogram of the Acceptance-Rejection samples follows the true density well. The simulated values have the expected parabolic shape. Small differences are due to random sampling variation.

### ii)

The empirical acceptance rate was `r round(empirical_rate, 4)`, which is close to the theoretical acceptance rate

\[
\frac{1}{M}
\approx0.6667.
\]

The small difference is due to random variation in the simulation.

---

### iii)
The inverse transform method is preferable when \(F^{-1}\) is available or canbe computed efficiently, since every generated uniform value produces one sample. The Acceptance-Rejection method is preferable when the inverse CDF is difficult to obtain but a good proposal distribution is available. Here, AR is reasonably efficient because its theoretical acceptance rate is \(2/3\).
